Network Visualizations in \MakeLowercase{\texttt{ggplot2}}

### 2015-10-11 inst/paper.Rnw
moving data statements into the code for individual examples

<<readingdata, eval=TRUE, echo=FALSE>>=
library(ggplot2) # needs to be version ≥ 1.0.1.9003
library(ggnet) # install from github
source("https://raw.githubusercontent.com/briatte/ggnet/master/ggnet2.R")
data(blood, football, theme_elements, lesmis,madmen, mm_directed, email,
     bikes)

@@

bloodnet <- merge(blood$edges, blood$vertices, by.x="from", by.y="label", all=TRUE)

employee <- data.frame(expand.grid(
              label=unique(email$nodes$label), day=unique(email$edges$day)))
employee <- merge(employee, email$nodes, by="label")
emailnet <- merge(subset(email$edges, nrecipients < 54), employee, by.x=c("From", "day"), by.y=c("label", "day"), all=TRUE)

@@

\section{Examples of Networks}
In this section, we demonstrate the current capabilities of \texttt{geom\_net} and \code{ggnet} in a series of side by side examples.
  \subsection{Blood Donation}
  In this directed network, there are eight vertices and 27 edges.  The vertices represent the eight different blood types in humans that are most important for donation: the ABO blood types A, B, AB, and O, combined with the RhD positive (+) and negative (-) types. The edges are directed: a person whose blood type is that of a \emph{from} vertex can to donate blood to a person whose blood type is that of a corresponding \emph{to} vertex. In the example below, loops are removed because loops exist on every vertex in this example, as blood between two people of matching ABO and RhD type can always be exchanged.
<<blood_ex, echo=TRUE, fig.cap= "Directed network of blood type donations in humans.", fig.width=6, fig.height=6>>=
ggplot(data = bloodnet, aes(from_id = from, to_id = to)) +
  geom_net(colour="darkred", layout="circle", label=TRUE, size=15,
           directed=TRUE, vjust=0.5, labelcolour="grey80",
           arrowsize=1.5, linewidth=0.75, arrowgap=0.05) + theme_net()
@

<<blood_ex_ggnet2, echo=TRUE, fig.cap= "Directed network of blood type donations in humans.", fig.width=6, fig.height=6>>=
ggnet2(network(blood$edges[, 1:2]), mode = "circle", size = 15,
       label = TRUE, arrow.size = 7.5, arrow.gap=0.05, vjust = 0.5,
       node.color="darkred", label.color="grey80")
@
This network is shown in figure \ref{fig.cap:blood_ex}.  Here, we have used the aesthetics \texttt{colour} and \texttt{size} set to identity values to change the size and color of all vertices. We have also used the \texttt{layout} and \texttt{label} arguments to change the default layout to a circle layout and to print the blood types, respectively. The circle layout places blood types of the same ABO type next to each other and spreads the vertices out far enough to distinguish between the various ``in" and ``out" types.  You can tell clearly from this plot that the O- type is the universal donor: it has an out-degree of seven and an in-degree of zero. Additionally, we can see that the AB+ type is the universal recepient, with an in-degree of seven and an out-degree of zero. Anyone looking at this plot can quickly determine which type(s) of blood they can receive and which type(s) can receive their blood.
\begin{figure}
\begin{subfigure}[t]{0.49\textwidth}
\caption{geom\_net}
\includegraphics[width=\textwidth]{figure/blood_ex-1.pdf}
\end{subfigure}
\begin{subfigure}[t]{0.49\textwidth}
\caption{ggnet2}
\includegraphics[width=\textwidth]{figure/blood_ex_ggnet2-1.pdf}
\end{subfigure}
\caption{\label{fig.cap:blood_ex} Network of blood donation possibilities in humans by ABO and RhD blood types.}
\end{figure}
\afterpage{\clearpage}

@@

<<email_ex, fig.cap = 'All emails within a company over a two week period.',echo=TRUE, out.width='\\textwidth'>>=
ggplot(data = emailnet, aes(from_id = From, to_id = to)) +
  geom_net(aes(colour= CurrentEmploymentType),
           linewidth=0.5, ealpha=0.5, size=4, directed=TRUE) +
  scale_colour_brewer("Employment Type", palette="Set1") + theme_net() +
  theme(legend.position="bottom")
@

@@

<<email_ex_facet, fig.height=5, fig.width=8, fig.cap = 'All emails within a company over a two week period facetted by day of the month.',echo=TRUE, out.width='\\textwidth'>>=
ggplot(data = emailnet, aes(from_id = From, to_id = to)) +
  geom_net(aes(colour= CurrentEmploymentType), fiteach=TRUE,
           linewidth=0.5, ealpha=0.1, size=2) +
  scale_colour_brewer(palette="Set1") + theme_net() +
  facet_wrap(~day, nrow = 2, labeller="label_both") +
  theme(legend.position="bottom",
        panel.border = element_rect(fill=NA, colour = "grey60"))
@

>>>>

<<readingdata, eval=TRUE, echo=FALSE>>=
library(ggplot2) # needs to be version ≥ 1.0.1.9003
library(ggnet) # install from github
source("https://raw.githubusercontent.com/briatte/ggnet/master/ggnet2.R")
data(football, theme_elements, lesmis,madmen, mm_directed, email,
     bikes)

@@

@@

\section{Examples of Networks}
In this section, we demonstrate the current capabilities of \texttt{geom\_net} and \code{ggnet} in a series of side by side examples.
  \subsection{Blood Donation}
  In this directed network, there are eight vertices and 27 edges.  The vertices represent the eight different blood types in humans that are most important for donation: the ABO blood types A, B, AB, and O, combined with the RhD positive (+) and negative (-) types. The edges are directed: a person whose blood type is that of a \emph{from} vertex can to donate blood to a person whose blood type is that of a corresponding \emph{to} vertex. In the example below, loops are removed because loops exist on every vertex in this example, as blood between two people of matching ABO and RhD type can always be exchanged.

<<blood_ex, echo=TRUE, fig.cap= "Directed network of blood type donations in humans.", fig.width=6, fig.height=6>>=
library(ggnet)
data(blood)

# source code for geom_net
ggplot(data = blood$edges, aes(from_id = from, to_id = to)) +
  geom_net(colour="darkred", layout="circle", label=TRUE, size=15,
           directed=TRUE, vjust=0.5, labelcolour="grey80",
           arrowsize=1.5, linewidth=0.75, arrowgap=0.05) + theme_net()
@

<<blood_ex_ggnet2, echo=TRUE, fig.cap= "Directed network of blood type donations in humans.", fig.width=6, fig.height=6>>=
# source code for ggnet2
library(GGally) # use developmental version on github for now

ggnet2(network(blood$edges[, 1:2]), mode = "circle", size = 15,
       label = TRUE, arrow.size = 7.5, arrow.gap=0.05, vjust = 0.5,
       node.color="darkred", label.color="grey80")
@
This network is shown in figure \ref{fig.cap:blood_ex}.  Here, we have used the aesthetics \texttt{colour} and \texttt{size} set to identity values to change the size and color of all vertices. We have also used the \texttt{layout} and \texttt{label} arguments to change the default layout to a circle layout and to print the blood types, respectively. The circle layout places blood types of the same ABO type next to each other and spreads the vertices out far enough to distinguish between the various ``in" and ``out" types.  You can tell clearly from this plot that the O- type is the universal donor: it has an out-degree of seven and an in-degree of zero. Additionally, we can see that the AB+ type is the universal recepient, with an in-degree of seven and an out-degree of zero. Anyone looking at this plot can quickly determine which type(s) of blood they can receive and which type(s) can receive their blood.
\begin{figure}
\begin{subfigure}[t]{0.49\textwidth}
\caption{geom\_net}
\includegraphics[width=\textwidth]{figure/blood_ex-1.pdf}
\end{subfigure}
\begin{subfigure}[t]{0.49\textwidth}
\caption{ggnet2}
\includegraphics[width=\textwidth]{figure/blood_ex_ggnet2-1.pdf}
\end{subfigure}
\caption{\label{fig.cap:blood_ex} Network of blood donation possibilities in humans by ABO and RhD blood types.}
\end{figure}
\afterpage{\clearpage}

@@

<<email_ex, fig.cap = 'All emails within a company over a two week period.',echo=TRUE, out.width='\\textwidth'>>=
emailnet <- merge(subset(email$edges, nrecipients < 54), email$nodes, by.x="From", by.y="label", all=TRUE)

ggplot(data = emailnet, aes(from_id = From, to_id = to)) +
  geom_net(aes(colour= CurrentEmploymentType),
           linewidth=0.5, ealpha=0.5, size=4, directed=TRUE) +
  scale_colour_brewer("Employment Type", palette="Set1") + theme_net() +
  theme(legend.position="bottom")
@

@@

<<email_ex_facet, fig.height=5, fig.width=8, fig.cap = 'All emails within a company over a two week period facetted by day of the month.',echo=TRUE, out.width='\\textwidth'>>=
employee <- data.frame(expand.grid(label=unique(email$nodes$label),
                                   day=unique(email$edges$day)))
employee <- merge(employee, email$nodes, by="label")
emailnet <- merge(subset(email$edges, nrecipients < 54), employee, by.x=c("From", "day"), by.y=c("label", "day"), all=TRUE)

ggplot(data = emailnet, aes(from_id = From, to_id = to)) +
  geom_net(aes(colour= CurrentEmploymentType), fiteach=TRUE,
           linewidth=0.5, ealpha=0.5, size=2, directed=TRUE,
           arrowsize=.5) +
  scale_colour_brewer(palette="Set1") + theme_net() +
  facet_wrap(~day, nrow = 2, labeller="label_both") +
  theme(legend.position="bottom",
        panel.border = element_rect(fill=NA, colour = "grey60"))
@
